---
title: Processing raw data from Vroomen et al. (2007)
author: Dave Kleinschmidt
date: "`r Sys.Date()`"
---

# Data organization and format

This directory contains the raw data as it came to me, in two batches.  The
processed data is taken from the `Data I` and `Data 13` subdirectories.  The
subject numbers for the `Data 13` batch need to be incremented by 16 (e.g.,
subject 1 in that subdirectory is 17 in the processed dataset) to avoid
conflicts with the `Data I` subject numbers.

There's one `.dat` data file for each block (two pretest blocks and a number of
adaptation blocks), with condition information embedded in the filenames.  I'm
using the _file's_ timestamps to get the order of the individual blocks (since
that varies across sessions).

Pre-test files have filenames starting with `Pre` or `pre`, and adaptation block
files have filenames starting with `v[bd]a[1-9]`, the name of the audio-visual
adaptation stimulus used on that block.

The individual files have the same format.  There are comment lines at the top
(lines starting with `!`), followed by a data table with one row per stimulus
and space separated columns for:

  * Trial (number)
  * Movie (filename)
  * Response ("Z" or "M", for "b" and "d" responses respectively, and "#" for no
    response)
  * Time (seconds with comma decimal. RT?).

# Ingesting raw data

```{r}
library(tidyverse)
library(magrittr)
library(forcats)
library(stringr)
library(assertthat)

files_pre <- Sys.glob("[dD]ata */[Pp][Rr][Ee]*")
files_adapt <- Sys.glob("[dD]ata */[Vv]*")

parse_response <- function(resp) {
  factor(resp, levels=c("Z", "M")) %>% fct_recode(b="Z", d="M")
}

read_file <- function(fn) {
  fn %>% 
    readr::read_table(
      col_names = c("trial", "movie", "response", "time"),
      col_types = "iccc",
      comment = "!"
    ) %>%
    ## manually parse time (decimal, sometimes with ',', sometimes with '.')
    mutate(time = time %>% stringr::str_replace(",", ".") %>% as.double(),
           source = fn,
           session_time = file.mtime(fn),
           source_lower = str_to_lower(source)) %>%
    tidyr::extract(source_lower, "subject",
                   regex="pp([0-9]*)",
                   remove=TRUE, convert=TRUE) %>%
    mutate(subject = ifelse(str_detect(source, "Data 13"), subject + 16, subject),
           response_raw = response,
           response = parse_response(response_raw))
}

```

# Pretest data

```{r}

dat_pre <-
  files_pre %>%
  map(read_file) %>%
  lift(bind_rows)()

pretest <- dat_pre %>%
  mutate(source_lower = str_to_lower(source)) %>%
  tidyr::extract(source_lower, "session", regex="pre([12])",
                 remove=TRUE, convert=TRUE) %>%
  tidyr::extract(movie, "aud", "bd([0-9]+)", convert=TRUE, remove=FALSE)

(pretest)
```

## Pretest sessions

There are two sessions: first with the whole continuum, second with just the
middle part that will be tested during recalibration

```{r}

ggplot(pretest, aes(x=aud)) +
  geom_histogram(breaks=seq(0.5,10)) +
  facet_grid(.~session)

```

## Category boundaries

```{r}

pretest %>%
  filter(session == 1) %>%
  ggplot(aes(x=aud, y=as.numeric(response=='b'), group=subject)) +
  geom_point(stat='summary', fun.y=mean) +
  stat_smooth(method='glm', method.args=list(family='binomial')) +
  facet_wrap(~subject)

```

# Adaptation data

```{r}

dat_adapt <-
  files_adapt %>%
  map(read_file) %>%
  lift(bind_rows)()

```

The adaptation data has exposure and test trials mixed together in the same data
files.  They're distinguished by their `movie` values: exposure movies are
audio-visual stimuli (starting with "v" or "V", followed by the visual category,
the audio stimulus number, and (if present) the number of repetitions).  The
test trials have the same filenames as in the pretest: "bd0[1-9]n.wav".  We need
to extract the number of repetitions for exposure trials, convert to cumulative
exposures, and combine with the test trials.  We also need to convert the
source/movie filenames into the condition variables (ambiguity and visual
category) and check for consistency.

## Source/movie consistency

```{r}

sessions <-
  dat_adapt %>%
  filter(str_detect(movie, "^[vV]")) %>%
  tidyr::extract(movie, c("movie_vid", "movie_aud"),
                 "[vV]([bd])a([0-9])", remove=FALSE, convert=TRUE) %>%
  tidyr::extract(source, c("source_vid", "source_aud"),
                 "[vV]([bd])a([0-9])", remove=FALSE, convert=TRUE)

mismatches <-
  sessions %>%
  group_by(source) %>%
  summarise_each(funs(unique), matches("(aud|vid)")) %>%
  filter((movie_vid != source_vid) | (movie_aud != source_aud))

mismatches %>% count(movie_aud, source_aud)
mismatches %>% count(movie_vid, source_vid)

```

There are quite a few blocks where the movie and source information mismatch.
Most of these are minor (the file says that `aud` is 5 but the movie is 4 or 6)
but some are major (a prototypical 1 or 9 instead of an ambiguous 4, 5 or 6, and
vice-versa).  I'm inclined to trust the movie field since that's formatted
consistently, leading me to think it was generated by the experiment script,
instead of the filenames which are much less consistent (and so probably done
manually).  But we'll want to check the counterbalancing and ordering, too.

```{r}

sessions <-
  sessions %>%
  group_by(source, movie_vid, movie_aud, subject, session_time) %>%
  summarise() %>%
  rename(exposure_vid = movie_vid, exposure_aud = movie_aud) %>%
  group_by(subject) %>%
  mutate(session = rank(session_time),
         exposure_vid = factor(exposure_vid),
         ambiguous = factor(ifelse(exposure_aud %in% c(1,9),
                                   "prototypical",
                                   "ambiguous"),
                            levels = c("prototypical", "ambiguous"))) %>%
  arrange(subject, session)

```

```{r}

sessions %>%
  ggplot(aes(x=session_time, y=subject)) +
  geom_point()

```

## Extracting exposure trials

The exposure and test trials are mixed together.  We need to know the cumulative
exposures for each test trial, so we need to pull out the repetitions on each
exposure movie and compute a cumulative sum.

```{r}

extract_exposures <- function(movie) {
  is_exposure <- str_detect(movie, "^[vV]")
  reps <- str_match(movie, "R([0-9]+)")[, 2] %>% as.double()
  ifelse(
    is_exposure,
    ifelse(
      is.na(reps),
      1,
      reps),
    0)
}

## add cumulative exposure counts, conditions, and session info to test trials
adapt <- 
  dat_adapt %>%
  mutate(exposures = extract_exposures(movie)) %>%
  group_by(source) %>%
  mutate(exposures = cumsum(exposures)) %>%
  # get rid of exposure trials
  filter(!str_detect(movie, "^[vV]")) %>%
  # pull out test stimulus continuum number
  tidyr::extract(movie, "aud", "bd([0-9]*)", convert=TRUE) %>%
  # combine with session metadata (condition, time, and session order)
  left_join(sessions) %>%
  ungroup()

```

# Save it

```{r}

pretest_mid <- pretest %>% filter(session == 2)
pretest <- pretest %>% filter(session == 1)

devtools::use_data(pretest, pretest_mid, adapt, overwrite=TRUE)

```
